---
title: "Anatomy of a Distribution"
author: "PJ Taal"
date: "2025-06-04"
output: html_document
---
---
title: "Anatomy of a Distribution"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(e1071)  # For skewness and kurtosis
```

## 1. Load and Inspect Data
We'll use the `mpg` dataset from the `ggplot2` package.

```{r load-data}
data <- mpg
summary(data$hwy)
```

## 2. Histogram
```{r histogram}
ggplot(data, aes(x = hwy)) +
  geom_histogram(bins = 15, fill = 'skyblue', color = 'black') +
  labs(title = "Histogram of Highway MPG", x = "Highway MPG", y = "Count")
```

## 3. Density Plot
```{r density-plot}
ggplot(data, aes(x = hwy)) +
  geom_density(fill = 'lightgreen') +
  labs(title = "Density Plot of Highway MPG")
```

## 4. Boxplot
```{r boxplot}
ggplot(data, aes(y = hwy)) +
  geom_boxplot(fill = 'orange') +
  labs(title = "Boxplot of Highway MPG", y = "Highway MPG")
```

## 5. Summary Statistics
```{r summary-stats}
hwy_stats <- data %>%
  summarise(
    mean = mean(hwy),
    median = median(hwy),
    sd = sd(hwy),
    skewness = skewness(hwy),
    kurtosis = kurtosis(hwy)
  )
hwy_stats
```

## 6. Q-Q Plot
Check normality visually.

A Quantile-Quantile plot is a graphical tool that is used to compare the quantiles of a dataset with the quantiles of a theoretical distribution (most often the Normal Distribution.)
   #If the data is from the theoretical distribution, points will lie along a straight 45 degree line. 

#Why use a Q-Q Plot?
   
   -check if it follows a normal distribution
   -detect skewness kurtosis and outliers
   -validate assumptions for parametric tests
   -compare two datasets. (will need more explanation for this)
```{r qqplot}
qqnorm(data$hwy)
qqline(data$hwy, col = "red")
```

## 7. Transformation Example
```{r log-transform}
data <- data %>% mutate(log_hwy = log(hwy))
ggplot(data, aes(x = log_hwy)) +
  geom_histogram(bins = 15, fill = 'purple', color = 'black') +
  labs(title = "Log-Transformed Histogram of Highway MPG")
```

## 8. Conclusion
This document explored the anatomy of a distribution using visualizations, summary statistics, and transformations to understand the shape, spread, and structure of a numeric variable.
