---
title: "dplyr exercise"
author: "PJ Taal"
date: "2025-06-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Library zoo
```{r}
library(dplyr)
library(tidyr)
```

Summarize cases

```{r}
mtcars |> summarize(avg=mean(mpg))

mtcars |> count(cyl)
```

Group Cases

```{r}
#Use group_by to create a "grouped copy of a table grouped by columns in ... dplyr functions will manipulate each "group" separately and combine the results.
mtcars |>
  group_by(cyl)|>
  summarize(avg = mean(mpg))
 
# Use Row wise to group data into individual rows. dplyr functions will compute results for each row. Also apply functions to list-columns. See tidyr cheat sheet for list-column workflow.  


mtcars |>
  group_by(cyl)|>
  summarize(avg = mean(mpg))


starwars |> 
  rowwise() |>
  mutate(film_count = length(films))

#ungroup returns ungrouped copy of table.
g_mtcars <- mtcars |>
  group_by(cyl) |>
  summarise(count=n())

ungroup(g_mtcars)
```


Manipulate Cases



## Extract Cases
    ### Row functions return subset of rows as a new table

```{r}
# filter | Extract rows that meet logical criteria
mtcars |> filter(mpg > 20)

# distinct | Remove rows with duplicate rows

```{r}
## Extract Cases

# filter | Extract rows that meet logical criteria
mtcars |> filter(mpg > 20)

# distinct | Remove rows with duplicate values

mtcars |> distinct(gear)

# slice | select row by position
mtcars |> slice(10:15)

# slice_sample | select row by position
mtcars |> slice_sample(n=5, replace= TRUE)

# slice_min | select rows with the lowest value, can use "n" number or proportion of lowest rows
mtcars |> slice_min(mpg,prop = 0.25)

#slice_max |  select rows with the highest value, can use "n" number or proportion of highest rows
mtcars |> slice_max(mpg,prop = 0.25)

#slice head / slice tail | select the first or last "n" number or proportion of rows
mtcars |> slice_head(n=5) 
mtcars |> slice_tail(prop = .15)

```


## Arrange Cases
    ### Row functions return rows by values of a column or columns ordered by high or low
    
```{r}
# arrange 
mtcars |> arrange(mpg)
mtcars |> arrange(desc(mpg))
```

## Add Cases
   ### Add a row to a table will and NA to unspecified value

```{r}
cars |> add_row()
```

Manipulate Variables

## Extract Variales
  ### Column functions return a set of columns as a new vector or table

```{r}
# pull | Extract column values as a vector, by name or index
mtcars |> pull(wt)

# select | Extract columns as a vector
mtcars |> select(mpg,wt)

# relocate
mtcars |> relocate(mpg, cyl, .after = last_col())

```

## MANIPULATE MULTIPLE VARIABLES AT ONCE
  ### 

```{r}
df <- tibble(x_1 = c(1, 2), x_2 = c(3, 4), y = c(4, 5))
```

```{r}
# across | Summarize or mutate multiple columns in the same way 
df |> summarize(across(everything(),mean))

#c_across
df |> 
  rowwise() |>
  mutate(x_total = sum(c_across(1:2)))
```

## MAKE NEW VARIABLES

```{r}
# Mutate
# Rename


```




